Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-16 manuscript/HIAs.qmd
add some refs

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
---

@@

Some general text about the decline in natural capital?

cite-style: super

Ecosystem condition accounting is the game of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (System of Environmental-Economic Accounting - Ecosystem Accounting).
The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.

Central to ecosystem condition accounts are variables and indicators.
These are parameters chosen to reflect the sentral condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
Similarly, a value of 0 mean that the variable is the worst possible state.
The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthorpogenic disturbance.

A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled (REF). This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition. SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e. not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator. There are at least 3 general ways to achieve this complete areal coverage of indicator values:

@@

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option b in the list above).
Models can describe the general associations between the spatially sampled data and the context (e.g. the set of environmental variables) where it was samples, and use these arelationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can be very reliable and make good indicators.
This is especially true when the ecosystem assets are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent lvel of uncertainty from such models becomes unacceptable.

@@

In this study we focused on open mire ecosystems in Norway.
For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model (REF Bakkestuen).
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire (Bakkestuen REF).
Mires are ecologically and socially important in Norway simply do to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment (REF Nybø, and web site).
This study builds on the work in that report.

@@

To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|---------------|---------------|---------------|---------------|---------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @rune_halvorsen_dokumentasjon_2019 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | ref                                |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | Mdir                               |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | Halvorsen                          |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | Mdir                               |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | Erikstad                           |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |

The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing (REF).
We included data fram 2018 to 2023.
In this survey, selected nature types are delineatied on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
Of all possible mire types, the survey only maps the following:

@@

7FK was developed into a single indicator (i.e. a normalised variable) named *Alien plant cover*.
Variables 2-5 describe very related aspects as so they where planned to be combined into one indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
This was done by taking the value from the variable with the lowest value, i.e. worst condition.
For both indicators, the lower and upper reference values, i.e. the worst and best possible condition that the variables can be in, is given by the range of the data where 0 or 1 is the lower limit and 4, 7 or 8 is the upper limit, depending on the variable (see @tbl-variables).

The Alien plant cover was assigned to ECT class B1 - Compositional state characteristics and ADSV was assigned to ECT class A1 - Physical state characteristics (REF).

@@

Schematic GIS workflow, see @fig-workflow

>>>>

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
---

@@

Ecosystem condition accounting is the game of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about.
A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.

Central to ecosystem condition accounts are variables and indicators.
These are parameters chosen to reflect the central condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
Similarly, a value of 0 mean that the variable is the worst possible state.
The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthropogenic disturbance.

A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets [@czucz_selection_2021, table 1; see also @unitednations2021 §2.87] Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled [@unitednations2021]. This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition. SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e. not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator. There are at least 3 general ways to achieve this complete areal coverage of indicator values:

@@

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option b in the list above).
Models can describe the general associations between the spatially sampled data and the context (e.g. the set of environmental variables) where it was samples, and use these relationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can be very reliable and make good indicators.
This is especially true when the ecosystem assets are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from such models becomes unacceptable.

@@

In this study we focused on open mire ecosystems in Norway.
For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment [@nybo_indikatorer_2023; see also @kolstad2023], and web site).
This study builds on the work in that report.

@@

To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).
The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing (REF).
We included data fram 2018 to 2023.
In this survey, selected nature types are delineatied on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
Of all possible mire types, the survey only maps the following:

@@

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|---------------|---------------|---------------|---------------|---------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | @runehalvorsen2019                 |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | @rune_halvorsen_dokumentasjon_2019 |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |

7FK was developed into a single indicator (i.e. a normalised variable) named *Alien plant cover*.
Variables 2-5 describe very related aspects and so they where combined into one indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
This was done by taking the value from the variable with the lowest value, i.e. worst condition.
For both indicators, the lower and upper reference values, i.e. the worst and best possible condition that the variables can be in, is given by the range of the data where 0 or 1 is the lower limit and 4, 7 or 8 is the upper limit, depending on the variable (see @tbl-variables).
See @fig-workflow for a schematic workflow for the indicator development.

The Alien plant cover was assigned to ECT class B1 - Compositional state characteristics, and ADSV was assigned to ECT class A1 - Physical state characteristics [@czucz_common_2021].

@@
